Developing an automated iterative near-term forecasting system for an ecological

### 2018-08-27 manuscript/portal_iterative_forecasting_ms.Rmd

---
title: Developing an automated iterative near-term forecasting system for an ecological
  study
author: Ethan P. White$^{1,2,3}$, Glenda M. Yenni$^1$, Shawn D. Taylor$^4$, Erica
  M. Christensen$^1$, Ellen K. Bledsoe$^4$, Juniper L. Simonis$^1$, S. K. Morgan Ernest$^{1,3}$
csl: mee.csl
output:
  pdf_document:
    keep_tex: yes
  html_document: default
geometry: left=3.5cm, right=3.5cm, top=2.25cm, bottom=2.25cm, headheight=12pt, letterpaper
header-includes:
- \usepackage{times}
- \usepackage{setspace}
- \usepackage{booktabs}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
fontsize: 12pt
bibliography: refs.bib
---
\raggedright

@@

1. Most forecasts for the future state of ecological systems are conducted once and never updated or assessed. As a result, many available ecological forecasts are not based on the most up-to-date data, and the scientific progress of ecological forecasting models is slowed by a lack of feedback on how well the forecasts perform.
2. Iterative near-term ecological forecasting involves repeated daily to annual scale forecasts of an ecological system as new data becomes available and regular assessment of the resulting forecasts. We demonstrate how automated iterative near-term forecasting systems for ecology can be constructed by building one to conduct monthly forecasts of rodent abundances at the Portal Project, a long-term study with over 40 years of monthly data. This system automates most aspects of the six stages of converting raw data into new forecasts: data collection, data sharing, data manipulation, modeling and forecasting, archiving, and presentation of the forecasts.
3. The forecasting system uses R code for working with data, fitting models, making forecasts, and archiving and presenting these forecasts. The resulting pipeline is automated using continuous integration (a software development tool) to run the entire pipeline once a week. The cyberinfrastructure is designed for long-term maintainability and to allow the easy addition of new models. Constructing this forecasting system required a team with expertise ranging from field site experience to software development.
4. Automated near-term iterative forecasting systems will allow the science of ecological forecasting to advance more rapidly and provide the most up-to-date forecasts possible for conservation and management. These forecasting systems will also accelerate basic science by allowing new models of natural systems to be quickly implemented and compared to existing models. Using existing technology, and teams with diverse skill sets, it is possible for ecologists to build these systems and use them to advance our understanding of natural systems.

@@

Forecasting the future state of ecological systems is important for management, conservation, and evaluation of how well models capture the processes governing ecological systems [@clark2001; @tallis2006; @diaz2015; @dietze2017]. In 2001, Clark et al. [-@clark2001] called for a more central role of forecasting in ecology. Since then, an increasing number of ecological forecasts are being published that focus on societally important questions from daily to decadal time scales [@dietze2018]. At daily scales, ecological forecasts predict the occurance of environmental issues like toxic algal blooms [@stumpf2009]. At quarterly scales, forecasts are used to predict the stocks of fisheries and make decisions about quotas for fishing [@NOAA2016]. At decadal time scales, ecological forecasts are used to predict how biodiversity will change as it responds to anthropogenic influences [@harris2018]. These forecasting examples highlight the important role that ecological forecasts play in recasting ecological knowledge in societally relevant ways and also improve our understanding of ecological systems by testing the ability of our models to predict how systems will change in the future [@dietze2018; @harris2018].

@@

Because iterative near-term forecasting requires a dynamic integration of models, predictions, and data, Dietze et al. [-@dietze2018] highlight approaches to data management, model construction and evaluation, and cyberinfrastructure that are necessary to effectively implement this type of forecasting (Box 1). Data needs to be released quickly under open licenses [@vargas2017; @dietze2018] and structured so that it can be used easily by a variety of researchers and in multiple modeling approaches [@borer2009; @strasser2011]. Models need to be able to deal with uncertainty, in both the predictors and the predictions, to properly convey uncertainty in the resulting forecasts [@diniz-filho2009]. Multiple models should be developed, both to assess which models are performing best [@dietze2018] and to facilitate combining models to form ensemble predictions which tend to perform better than single models [@araujo2007; @diniz-filho2009]. Ensuring that data and models are regularly updated and new forecasts are made requires cyberinfrastructure to automate data processing, model fitting, prediction, model evaluation, forecast visualization, and archiving. In combination, these approaches should allow forecasts to be easily rerun and evaluated as new data becomes available [Box 1; @dietze2018].

@@

Iterative forecasting is most effective with frequently collected data, since it provides more opportunities for updating model results and assessing (and potentially improving) model performance [Box 1; @dietze2018]. The Portal Project is a long-term ecological study situated in the Chihuahuan Desert (2 km north and 6.5 km east of Portal, Arizona, US). Researchers have been continuously collecting data at the site since 1977, including data on the abundance of rodent and plant species (monthly and twice yearly, respectively) and climatic factors such as air temperature and precipitation (daily) [@brown1998; @ernest2009; @ernest2016; @ernest2018]. The site consists of 24 50m x 50m experimental plots. Each plot contains 49 permanently marked trapping stations laid out in a 7 x 7 grid, and all plots are trapped with Sherman live traps for one night each month. For all rodents caught during a trapping session, information on species identity, size, and reproductive condition is collected, and new individuals are given identification tags. This information on rodent populations is high-frequency, uses consistent trapping methodology, and has an extended time-series (475 monthly samples and counting), making this study an ideal case for near-term iterative forecasting.

@@

Implementation of iterative forecasting requires the regular rebuilding of models with new raw data as it becomes available and the presentation of those forecasts in usable forms; in our case, this occurs monthly. Rebuilding models in an efficient and maintainable way relies on developing an automated pipeline to handle the six stages of converting raw data into new forecasts: data collection, data sharing, data manipulation, modeling and forecasting, archiving, and presention of the forecasts (Figure 1a). To implement the pipeline outlined in Figure 1a, we used a "continuous analysis" framework [*sensu* @beaulieu2017] that automatically processes the most up-to-date data, refits the models, makes new forecasts, archives the forecasts, and updates a website with analysis of current and previous forecasts. In this section we describe our approach to streamlining and automating the multiple components of the forecasting pipeline and the tools and infrastructure we employed to execute each component.

@@

In addition to automatically running software pipelines, the other key component of "continuous analysis" is making sure that the pipelines will continue to run even as software dependencies change [@beaulieu2017]. Many of us have experienced the frustrations that can occur when software updates (e.g., changes in R package versions) create errors in previously functional code. We experienced this issue when the `tscount` package [@liboschik2015], used by two of our forecasting models, was temporarily removed from CRAN (the R package repository) and could not be installed in the usual way. This broke our forecasting pipeline, as we could no longer run models that used that package. To make our pipeline robust to changes in external software dependencies, we follow Beaulieu and Greene's [-@beaulieu2017] recommendation to use software containers. Software containers are standalone packages that contain copies of everything needed to run a given piece of software, including the operating system [@boettiger2015]. Once created, a software container is basically a time capsule, containing all the software dependencies in the exact state used to develop and run the software [@boettiger2015]. If those dependencies change (or disappear) in the wider world, they still exist, unchanged, in the container. We use an existing platform, Docker [@merkel2014], to store an exact image of the complete software environment for running the forecasts. Docker also allows a specified set of packages to be used consistently across different computer and server environments. Using containers allows us to control transitions to new package versions, implementing them only after we have tested them and made any necessary changes to the data processing and analysis code. We use a container created by the Rocker project, which is a Docker image with many important R packages [i.e., the tidyverse packages; @tidyverse] pre-installed  [@boettiger2017]. We add our code and dependencies to this existing Rocker image to create a software container for our forecasting pipeline. In combination, the automated running of the pipeline (continuous integration) and the guarantee it will not stop working unexpectedly due to software dependencies (via a software container) allows continuous analysis to serve as the glue that connects all stages of the forecasting pipeline.

@@

Iterative near-term forecasting involves regularly refitting a variety of different models (Figure 1). Ideally, new models should be easy to incorporate to allow for iterative improvements to the general modeling structure and approach. We use CI to refit the models and make new forecasts each time the modeling code changes and when new data become available (Figure 1b). We use a plugin infrastructure to allow new models to be easily added to the system. This approach treats each model as an interchangable black box; all models have access to the same input data and generate the same structure for model outputs (Figure 2). During each run of the forecasting code, all existing models are run and the standardized outputs are combined into a single file to store the results of the different models' forecasts. A weighted ensemble model is then added with weights based on how well individual models fit the training data. This plugin infrastructure makes it easy to add and compare very different types of models, from the basic time-series approaches currently implemented to the more complex state-space and machine learning models we hope to implement in the future. As long as a model script can load the provided data and produce the appropriate output, it will be run and its results incorporated into the rest of the forecasting system. This means that anyone can add a new model to the existing system by: 1) creating their own copy of the project (typically by forking the project on GitHub); 2) developing a new model; and 3) submitting a pull request to our repository.

@@

Forecasts are currently evaluated using root mean square error (RMSE) to evaluate point forecasts and coverage to evaluate uncertainty. We plan to add additional metrics in the future. In addition to evaluating the actual forecasts, we also use hindcasting [forecasting on already collected data; @jolliffe2003] to gain additional insight into the methods that work best for forecasting this system. For example, a model is fit using rodent observations up to June 2005, then used to make a forecast 12 months out to May 2006. The observations of that 12-month period can immediately be used to evaluate the model. Since hindcasting is conducted using data that has already been collected, it allows model comparisons to be conducted on large numbers of hindcasts and provides insight into which models make the best forecasts without needing to wait for new data to be collected [@harris2018]. It can also be used to quickly evaluate new models instead of waiting for an adequate amount of data to accumulate. As the performance of different models is understood through evaluation of forecasts and hindcasts, models can be refined or removed from the system or ensemble to iteratively improve the resulting forecasts.

@@

Following the recommendations of Dietze et al [-@dietze2018], we developed an automated iterative forecasting system (Figure 1) to support repeated forecasting of an ecological system. Our forecasting system automatically acquires and processes the newest data, refits the models, makes new forecasts, publicly archives those forecasts, and presents both the current forecast and information on how previous forecasts performed. Every week, the forecasting system generates a new set of forecasts with no human intervention, except for the entry of new field data. Our forecasting system ensures that forecasts based on the most recent data are always available and is designed to allow rapid assessment of the performance of multiple forecasting models for a number of different states of the system, including the abundances of individual species and community-level variables such as total abundance. To create this iterative near-term forecasting system, we used R to process data and conduct analyses and leveraged existing tools and services (i.e. GitHub, Travis, Docker) for more complicated cyberinfrastructure tasks. Thus, our approach to developing iterative near-term forecasting infrastructure provides an example for how short-term ecological forecasting systems can be developed.

@@

@@

@@

Understanding the uncertainty of forecasts is crucial to interpreting and understanding their utility. Models used for forecasting should be probabilistic to properly quantify uncertainty and to convey how this uncertainty increases through time. Evaluation of forecast models should include assessment of how accurately they quantify uncertainty as well as point estimates  [@hooten2015].

@@

Each forecast iteration involves acquiring new data, refitting the models, and making new forecasts. This should be done automatically without requiring human intervention. Therefore, the process of making forecasts should emphasize end-to-end reproducibility, including data, models, and evaluation [@stodden2014], to allow the forecasts to be easily rerun as new data becomes available [@dietze2018].

>>>>

---
title: Developing an automated iterative near-term forecasting system for an ecological
  study
author: Ethan P. White$^{1,2,3}$, Glenda M. Yenni$^1$, Shawn D. Taylor$^4$, Erica
  M. Christensen$^1$, Ellen K. Bledsoe$^4$, Juniper L. Simonis$^1$, S. K. Morgan Ernest$^{1,3}$
csl: mee.csl
output:
  html_document: default
  pdf_document:
    keep_tex: yes
geometry: left=3.5cm, right=3.5cm, top=2.25cm, bottom=2.25cm, headheight=12pt, letterpaper
header-includes:
- \usepackage{times}
- \usepackage{setspace}
- \usepackage{booktabs}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
fontsize: 12pt
bibliography: refs.bib
---
\raggedright

@@

1. Most forecasts for the future state of ecological systems are conducted once and never updated or assessed. As a result, many available ecological forecasts are not based on the most up-to-date data, and the scientific progress of ecological forecasting models is slowed by a lack of feedback on how well the forecasts perform.
2. Iterative near-term ecological forecasting involves repeated daily to annual scale forecasts of an ecological system as new data becomes available and regular assessment of the resulting forecasts. We demonstrate how automated iterative near-term forecasting systems for ecology can be constructed by building one to conduct monthly forecasts of rodent abundances at the Portal Project, a long-term study with over 40 years of monthly data. This system automates most aspects of the six stages of converting raw data into new forecasts: data collection, data sharing, data manipulation, modeling and forecasting, archiving, and presentation of the forecasts.
3. The forecasting system uses R code for working with data, fitting models, making forecasts, and archiving and presenting these forecasts. The resulting pipeline is automated using continuous integration (a software development tool) to run the entire pipeline once a week. The cyberinfrastructure is designed for long-term maintainability and to allow the easy addition of new models. Constructing this forecasting system required a team with expertise ranging from field site experience to software development.
4. Automated near-term iterative forecasting systems will allow the science of ecological forecasting to advance more rapidly and provide the most up-to-date forecasts possible for conservation and management. These forecasting systems will also accelerate basic science by allowing new models of natural systems to be quickly implemented and compared to existing models. Using existing technology, and teams with diverse skill sets, it is possible for ecologists to build automated forecasting systems and use them to advance our understanding of natural systems.

@@

Forecasting the future state of ecological systems is important for management, conservation, and evaluation of how well models capture the processes governing ecological systems [@clark2001; @tallis2006; @diaz2015; @dietze2017]. In 2001, Clark et al. [-@clark2001] called for a more central role of forecasting in ecology. Since then, an increasing number of ecological forecasts are being published that focus on societally important questions from daily to decadal time scales [@dietze2018]. At daily scales, ecological forecasts predict the occurrence of environmental issues like toxic algal blooms [@stumpf2009] and pollen [@prank2013]. At monthly scales, forecasts are used to predict the stocks of fisheries [@NOAA2016] and the probability of coral bleaching events [@liu2018]. At decadal time scales, ecological forecasts are used to predict how biodiversity will change as it responds to anthropogenic influences [@harris2018]. These forecasting examples highlight the important role that ecological forecasts play in recasting ecological knowledge in societally relevant ways and also improve our understanding of ecological systems by testing the ability of our models to predict how systems will change in the future [@dietze2018; @harris2018].

@@

Because iterative near-term forecasting requires a dynamic integration of models, predictions, and data, Dietze et al. [-@dietze2018] highlight approaches to data management, model construction and evaluation, and cyberinfrastructure that are necessary to effectively implement this type of forecasting (Box 1). Data needs to be released quickly under open licenses [@vargas2017; @dietze2018] and structured so that it can be used easily by a variety of researchers and in multiple modeling approaches [@borer2009; @strasser2011]. Models need to be able to deal with uncertainty, in both the predictors and the predictions, to properly convey uncertainty in the resulting forecasts [@diniz-filho2009]. Multiple models should be developed, both to assess which models are performing best [@dietze2018] and to facilitate combining models to form ensemble predictions which tend to perform better than single models [@araujo2007; @diniz-filho2009; @dormann2018]. Ensuring that data and models are regularly updated and new forecasts are made requires cyberinfrastructure to automate data processing, model fitting, prediction, model evaluation, forecast visualization, and archiving. In combination, these approaches should allow forecasts to be easily rerun and evaluated as new data becomes available [Box 1; @dietze2018].

@@

Iterative forecasting is most effective with frequently collected data, since it provides more opportunities for updating model results and assessing (and potentially improving) model performance [Box 1; @dietze2018]. The Portal Project is a long-term ecological study situated in the Chihuahuan Desert (2 km north and 6.5 km east of Portal, Arizona, US). Researchers have been continuously collecting data at the site since 1977, including data on the abundance of rodent and plant species (monthly and twice yearly, respectively) and climatic factors such as air temperature and precipitation (daily) [@brown1998; @ernest2009; @ernest2016; @ernest2018]. The site consists of 24 50m x 50m experimental plots. Each plot contains 49 permanently marked trapping stations laid out in a 7 x 7 grid, and all plots are trapped with Sherman live traps for one night each month. For all rodents caught during a trapping session, information on species identity, size, and reproductive condition is collected, and new individuals are given identification tags. This information on rodent populations is high-frequency, uses consistent trapping methodology, and has an extended time-series (475 monthly samples and counting), making this study an ideal case for near-term iterative forecasting. Forecasting of rodent population dynamics in the southwest (and more broadly) is important because of their link to zoonotic diseases such as hantavirus and plague [@parmenter1993; @gage2005; @springer2016]. In addition, this forecasting system is being used to improve population dynamic modeling for this community and to explore the utility of incorporating experimental data into ecological forecasting models.

@@

Implementation of iterative forecasting requires the regular updating of models with new raw data as it becomes available and the presentation of those forecasts in usable forms; in our case, this occurs monthly. Updating models in an efficient and maintainable way relies on developing an automated pipeline to handle the six stages of converting raw data into new forecasts: data collection, data sharing, data manipulation, modeling and forecasting, archiving, and presention of the forecasts (Figure 1a). To implement the pipeline outlined in Figure 1a, we used a "continuous analysis" framework [*sensu* @beaulieu2017] that automatically processes the most up-to-date data, updates the models, makes new forecasts, archives the forecasts, and updates a website with analysis of current and previous forecasts. In this section we describe our approach to streamlining and automating the multiple components of the forecasting pipeline and the tools and infrastructure we employed to execute each component.

@@

To ensure that software pipelines continue to run automatically as software dependencies change, a key component of "continuous analysis" is the use of a reproducible computational environment [@beaulieu2017]. We followed Beaulieu and Greene's [-@beaulieu2017] recommendation to use software containers. Software containers are standalone packages that contain copies of everything needed to run a given piece of software, including the operating system [@boettiger2015]. Once created, a software container is basically a time capsule, containing all the software dependencies in the exact state used to develop and run the software [@boettiger2015]. If those dependencies change (or disappear) in the wider world, they still exist, unchanged, in the container. We use an existing platform, Docker [@merkel2014], to store an exact image of our complete software environment by adding our project specific code to a container created by the Rocker project, which is a Docker image with many important R packages [i.e., the tidyverse packages; @tidyverse] pre-installed  [@boettiger2017]. We implemented this system because we experienced issues with external dependencies breaking our pipeline (e.g., when the `tscount` package [@liboschik2015], was temporarily removed from CRAN and could not be installed in the usual way). In combination, the automated running of the pipeline (continuous integration) and the guarantee it will not stop working unexpectedly due to software dependencies (via a software container) allows continuous analysis to serve as the glue that connects all stages of the forecasting pipeline.

@@

Iterative near-term forecasting involves regularly updating a variety of different models (Figure 1). Ideally, new models should be easy to incorporate to allow for iterative improvements to the general modeling structure and approach. We use CI to update the models and make new forecasts each time the modeling code changes and when new data become available (Figure 1b). We use a modular plugin infrastructure to allow new models to be easily added to the system. This approach treats each model as an interchangable black box; all models have access to the same input data and generate the same structure for model outputs (Figure 2). Details of how to add a new model to the system are provided in the core GitHub repository (https://github.com/weecology/portalPredictions/wiki/Adding-a-new-model). During each run of the forecasting code, all existing models are run and the standardized outputs are combined into a single file to store the results of the different models' forecasts. A weighted ensemble model is then added with weights based on how well individual models fit the training data. This plugin infrastructure makes it easy to add and compare very different types of models, from the basic time-series approaches currently implemented to the more complex state-space and machine learning models we hope to implement in the future. As long as a model script can load the provided data and produce the appropriate output, it will be run and its results incorporated into the rest of the forecasting system. This means that anyone can add a new model to the existing system by: 1) creating their own copy of the project (typically by forking the project on GitHub); 2) developing a new model; and 3) submitting a pull request to our repository.

@@

Forecasts are currently evaluated using root mean square error (RMSE) to evaluate point forecasts and coverage to evaluate uncertainty. We plan to add additional metrics, like deviance, that incorporate both accuracy and uncertainty and better match the calibration method [@hooten2015; @dietze2018]. In addition to evaluating the actual forecasts, we also use hindcasting [forecasting on already collected data; @jolliffe2003] to gain additional insight into the methods that work best for forecasting this system. For example, a model is fit using rodent observations up to June 2005, then used to make a forecast 12 months out to May 2006. The observations of that 12-month period can immediately be used to evaluate the model. Since hindcasting is conducted using data that has already been collected, it allows model comparisons to be conducted on large numbers of hindcasts and provides insight into which models make the best forecasts without needing to wait for new data to be collected [@harris2018]. It can also be used to quickly evaluate new models instead of waiting for an adequate amount of data to accumulate. As the performance of different models is understood through evaluation of forecasts and hindcasts, models can be refined or removed from the system or ensemble to iteratively improve the resulting forecasts.

@@

Following the recommendations of Dietze et al [-@dietze2018], we developed an automated iterative forecasting system (Figure 1) to support repeated forecasting of an ecological system. Our forecasting system automatically acquires and processes the newest data, updates the models, makes new forecasts, publicly archives those forecasts, and presents both the current forecast and information on how previous forecasts performed. Every week, the forecasting system generates a new set of forecasts with no human intervention, except for the entry of new field data. Our forecasting system ensures that forecasts based on the most recent data are always available and is designed to allow rapid assessment of the performance of multiple forecasting models for a number of different states of the system, including the abundances of individual species and community-level variables such as total abundance. To create this iterative near-term forecasting system, we used R to process data and conduct analyses and leveraged existing tools and services (i.e. GitHub, Travis, Docker) for more complicated cyberinfrastructure tasks. Thus, our approach to developing iterative near-term forecasting infrastructure provides an example for how short-term ecological forecasting systems can be developed.

@@

Expanding the community of ecological forecasters using continous analysis approaches will require both an expansion of the current toolkit and the development of standards to facilitate interoperability of forecasts and forecasting systems. One of the major challenges for our current forecasting system is supporting computationally intensive forecasts. Projects involving larger datasets and/or complex modelling approaches will require either hosted solutions that provide infrastructure for running continuous integration and allow long-running distributed jobs or solutions that involve the user setting up their own continuous analysis sytem on cloud infrastructure or high performance computing centers. Event-driven serverless cloud platforms like OpenWhisk (https://openwhisk.apache.org/) and AWS Lambda (https://aws.amazon.com/lambda/) provide potential as hosted solutions, and open source continuous integrations systems like Jenkins (https://jenkins.io/) can be integrated with either cloud or high performance computing centers. However, both solutions are currently more complicated to set up than the hosted continuous integration approach we have employed using Travis. In addition to scalability issue for more computationally intensive projects, the toolkit for continuous analysis needs be made more researcher friendly. To broaden the user-base that can use continuous analysis for forecasting, we recommend the development of tools that make setting up continuous analysis easier by automating configuration steps. We also recommend the development of tools or data repository infrastructure to support the easy automated archiving of regularly generated data and forecasts [see @yenni2018]. Finally, the development of standards for ecological forecasting to allow interoperability among forecasting systems will be essential for the growth of this field [see discussions of data standards, meta-data, and ontologies in ecology more broadly @jones2006; @madin2008; @michener2012]. Now is an opportune time for developing these standards while the community of ecological forecasters is still small. While we have developed an initial format for storing and sharing forecasts, it is still lacking in several areas. Most notably, our approach to storing information on models and their associated uncertainty is insufficient for all but the simplest models. Improving this framework will require capturing both covariances between state variables and the full uncertainty in the models, by either storing full model objects or additional information like full ensembles of predictions (e.g., from Monte Carlo based approaches). This is challenging due to a lack of general standards for reporting uncertainty [@dietze2018].

@@

# Author Contributions

All authors conceived the ideas and designed methodology; All authors developed the automated forecasting system; EPW and SKME led the writing of the manuscript. All authors contributed critically to the drafts and gave final approval for publication.

@@

Understanding the uncertainty of forecasts is crucial to interpreting and understanding their utility. Models used for forecasting should be probabilistic to properly quantify uncertainty and to convey how this uncertainty increases through time. Evaluation of forecast models should include assessment of how accurately they quantify uncertainty as well as point estimates  [@hooten2015; @harris2018].

@@

Each forecast iteration involves acquiring new data, updating the models, and making new forecasts. This should be done automatically without requiring human intervention. Therefore, the process of making forecasts should emphasize end-to-end reproducibility, including data, models, and evaluation [@stodden2014], to allow the forecasts to be easily rerun as new data becomes available [@dietze2018].
